---
title: "Week 4"
author: "Me"
date: "12/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#clustering case study - Understanding human activities with samsung

- Have a sense of what you are looking for
- produce a rough cut of analysis
- rules out certain questions.


```{r}
load('samsungData.rda')
names(samsungData)[1:12]
```
```{r}
table(samsungData$activity)
```

Separate out 6 activities based on accelerometer.

Plotting average acceleration for the first subject
```{r}
par(mfrow = c(1,2), mar = c(5,4,1,1))

samsungData = transform(samsungData, activity = factor(activity))

sub1 = subset(samsungData, subject == 1)


plot(sub1[,1], col = sub1$activity, ylab = names(sub1)[1])
plot(sub1[,2], col = sub1$activity, ylab = names(sub1)[2])
legend('bottomright', legend = unique(sub1$activity), col = unique(sub1$activity), pch = 1)

```

```{r}
source('plclust.R')
distance_matrix = dist(sub1[,1:3])
hclustering = hclust(distance_matrix)
myplclust(hclustering, lab.col =  unclass(sub1$activity))
```
No info from here.

Plotting max acceleration
```{r}
par(mfrow = c(1,2))
plot(sub1[,10], pch = 19, col = sub1$activity, ylab=  names(sub1)[10])
plot(sub1[,11], pch = 19, col = sub1$activity, ylab=  names(sub1)[11])
```


```{r}
distance_matrix = dist(sub1[,10:12])
hclustering = hclust(distance_matrix)
myplclust(hclustering, lab.col =  unclass(sub1$activity))
```
There's clear distinction between moving and non moving but on the left you can see that the activities are jumbled. Cols 10 and 11 can separate moving and non moving.

Use singular value decomposition
```{r}
svd1 = svd(scale(sub1[,-c(562,563)]))

par(mfrow = c(1,2))
plot(svd1$u[,1], col = sub1$activity, pch = 19)
plot(svd1$u[,2], col = sub1$activity, pch = 19)
```
First singular vector separates moving and non moving. Second singular vector is more vague. 

Find maximum contributor
```{r}
plot(svd1$v[,2], pch = 19)
```

added max contribution 
```{r}
maxContrib = which.max(svd1$v[,2])
distance_matrix = dist(sub1[,c(10:12, maxContrib)])
hclustering = hclust(distance_matrix)
myplclust(hclustering, lab.col =  unclass(sub1$activity))

```
More separated moving activities. Non moving activities are not yet separated. 

```{r}
names(samsungData)[maxContrib]
```

Kmeans clustering(nstart = 1)
```{r}
kclust = kmeans(sub1[, -c(562,563)], centers = 6)
table(kclust$cluster, sub1$activity)
```

Kmeans had trouble sepearting laying sitting and standing. 

Use 100 starting points
```{r}
kclust = kmeans(sub1[, -c(562,563)], centers = 6, nstart = 100)
table(kclust$cluster, sub1$activity)
```

Each cluster has a mean value or center. We can see which features seem to drive the location of center.

Cluster 1 (laying)
```{r}
plot(kclust$centers[1, 1:10], pch= 19, ylab = 'Cluster Center', xlab = '' )
```

Cluster walking
```{r}
plot(kclust$centers[4,1:10], pch= 19, ylab = 'Cluster Center', xlab = '' )
```

Which features have interesting values that drives that classification. 

#Case study - Air pollution

Is the air pollution in 2012 lesser than 1999
```{r}
library(rio)
library(tidyverse)
library(data.table)

nine = fread('https://raw.githubusercontent.com/jtleek/modules/master/04_ExploratoryAnalysis/CaseStudy/pm25_data/RD_501_88101_1999-0.txt') %>%
      janitor::clean_names()

head(nine)
```

Get pm2.5
```{r}
x0 = nine$number_of_samples
class(x0)
str(x0)
summary(x0)
mean(is.na(x0))
```

2012 data
```{r}
pm1 = fread('https://raw.githubusercontent.com/jtleek/modules/master/04_ExploratoryAnalysis/CaseStudy/pm25_data/RD_501_88101_2012-0.txt') %>% janitor::clean_names()

dim(pm1)

x1 = pm1$number_of_samples
summary(x1)
mean(is.na(x1))

boxplot(x0, x1)
```
Both are right skewed

```{r}
boxplot(log(x0), log(x1))
```

Why are there negative values?
```{r}
library(tidyverse)
negative = x1 < 0

sum(negative, na.rm = T)
mean(negative, na.rm = T)

dates = pm1$year
str(dates)

#convert dates
dates = as.Date(as.character(dates), '%Y%m%d')

str(dates)
hist(dates, 'month')

#dates with negative values
hist(dates[negative], 'month')
```
Exploring change at one monitor

```{r}
site0 = unique(subset(nine, state_code == 36, c('county_code','site_id')))
site1 = unique(subset(pm1, state_code == 36, c('county_code','site_id')))

site0 = paste(site0[,1],site0[,2], sep = '.')
site1 = paste(site1[,1],site1[,2], sep = '.')
```

```{r}
nine %>%
      filter(state_code == 36) %>%
      unique() %>%
      select(county_code, site_id)
```

Cant follow lecture because the dataset is different.




















