---
title: "Week 1 notes"
author: "Me"
date: "12/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(datasets)
library(tidyverse)
pollution = read_csv('avgpm25.csv') %>%
      mutate(region = factor(region))
```

```{r}
boxplot(pm25 ~region, data = pollution, col = 'red')
```

Multiple histograms
```{r}
par(mfrow = c(2,1), mar = c(4,4,2,1))
hist(subset(pollution, region == 'east')$pm25, col = 'green')
hist(subset(pollution, region == 'west')$pm25, col = 'green')
```

Scatterplot
```{r}
with(pollution, plot(latitude, pm25, col = region))
abline(h = 12, lwd = 2, lty= 2)
```
Multiple scatterplot
```{r}
par(mfrow = c(1,2), mar = c(5,4,2,1))
with(subset(pollution, region == 'west'), plot(latitude, pm25, main = 'west'))
with(subset(pollution, region == 'east'), plot(latitude, pm25, main = 'east'))
```

#Plotting systems in R
Base plot
```{r}
data(cars)
with(cars, plot(speed, dist))
```
```{r}
library(lattice)
state = data.frame(state.x77, region = state.region)
xyplot(Life.Exp ~ Income | region, data = state, layout = c(4,1))
```
```{r}
library(ggplot2)
data(mpg)
qplot(displ, hwy, data = mpg)
```


##Base plotting system

graphics: contains plotting functions for the base graphing systems including plot, hist, boxplot etc
grDevices: contains all code implementing various graphics devices including X11, PDF, postscript, PNG, etc

Lattice system is encapsulated in lattice package. 
lattice: contains code for producing trellis graphics, which are independent of the 'base' graphics system, includef functions like xyplot, bwplot, levelplot
grid: implements different graphing system independent of the 'base' system, the lattice package builds on top of grid. Seldom used directly.


Base graphics
- two phase in creating plot
  - initialize a new plot
  - annotating to an existing plot
- plot or hist will launch graphics device and draw a plot on device
- if the arguments in plot are not special class, then default method for plot is called. 
- the base graphics has many parameters documented in ?par


```{r}
hist(airquality$Ozone) #draws new plot
```
```{r}
with(airquality, plot(Wind, Ozone))
```

```{r}
airquality = transform(airquality, Month = factor(Month))
boxplot(Ozone ~ Month, airquality, xlab = 'Month', ylab = 'Ozone (pb)')
```

Other key parameters
- pch: plotting symbol
- lty : line type
- lwd : line width
- col : color. use colors() for vector of names
- xlab : x-axis label
- ylab : y-axis label

Some important paramets
par() - function is used to specify global graphics parameter that affect all plots in an R session
- las : orientation of the axis label
- bg : background color
- mar : margin size
- oma : outer margin size
- mfrow : number of plots per row, column (plots are filled row wise)
- mfcol : number of plots per row, column (plots are filled col wise)

Default values
```{r}
par('lty')
par('col')
par('pch')
par('bg')
par('mar') #4 sides down, left, up, right
par('mfrow')
```

Base plotting functions
- plot : makes a scatterplot, or other types depending on the class of object
- lines : add lines to plot. given a vector of x values and corresponding y or 2 column matrix
- points : add points
- text : add text labels using specified x,y coords
- title : add annotations outside the plot like title, subtitle, outer margine
- mtext : place text in the margins
- axis : axis ticks

```{r}
with(airquality, plot(Wind, Ozone))
title(main = 'Ozone and Wind in New York') # add title
```

```{r}
with(airquality, plot(Wind, Ozone, main = 'Ozone and Wind in New York'))
with(subset(airquality, Month == 5), points(Wind, Ozone, col = 'blue')) #add blue colors
```

With annotation
```{r}
with(airquality, plot(Wind, Ozone, main = 'Ozone and Wind in New York City', type = 'n')) #does not plot anything
with(subset(airquality, Month == 5), points(Wind, Ozone, col = 'blue'))
with(subset(airquality, Month != 5), points(Wind, Ozone, col = 'red'))
legend('topright', pch = 1, col = c('blue', 'red'), legend = c('May', 'Other months'))
```

Base plot with regression line
```{r}
with(airquality, plot(Wind, Ozone, main = 'Ozone and Wind in New York City', pch = 20))
model = lm(Ozone ~ Wind, airquality)
abline(model, lwd = 1)
```

Multiple baseplots
```{r}
par(mfrow = c(1,2))
with(airquality, {
      plot(Wind, Ozone, main = 'Ozone and Wind')
      plot(Solar.R, Ozone, main = 'Ozone and Solar Rad')
})
```

Multiple baseplots
```{r}
par(mfrow = c(1,3), mar = c(4,4,2,1), oma = c(0, 0, 2,0))
with(airquality, {
      plot(Wind, Ozone, main = 'Ozone and Wind')
      plot(Solar.R, Ozone, main = 'Ozone and Solar Rad')
      plot(Temp, Ozone, main = 'Ozone and Temp')
      mtext('Ozone and weather in New York city', outer = T)
})
```
##Base plotting demo
```{r}
x = rnorm(100)
hist(x)
```

```{r}
y = rnorm(100)
plot(x,y)
```

```{r}
par(mar = c(4,4,2,2))
plot(x,y)
```

check different points
```{r}
example(points)
```


```{r}
x = rnorm(100)
y = rnorm(100)
plot(x,y)
title('Scatterplot')
text(-2,-2, 'Label')
fit = lm(y ~ x)
abline(fit)
```

#Graphics devices in R
```{r}
?Devices
```

How does a plot get created?
```{r}
library(datasets)
with(faithful, plot(eruptions, waiting))
title(main = 'Old faithful Geyser data')
```
Using graphic device
```{r}
pdf(file = 'myplot.pdf')

#ccreate plot and send to file ( no plot on screen)
with(faithful, plot(eruptions, waiting))
title(main = 'Old faithful Geyser data')

#close pdf file
dev.off()
```


```{r}
dev.cur()
```














