---
title: "Week 2"
author: "Me"
date: "12/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Lattice plotting system
- for multiple plots
- calls from grid package
- create entire plot at once in a single function call.

Lattice functions:
- xyplot : scatterplots
- bwplot : boxpots
- histogram : histograms
- stripplot : boxplot that are points
- dotplot : violin strings
- splom : scatterplot matrix like pairs
- levelplot, contourplot for image data.  

Xyplot
- takes formula
- y ~  x | f * g, data
- f and g are categorical that you condition on
- scatterplot of y and x for every level of f and g

```{r}
library(lattice)
library(datasets)

xyplot(Ozone ~ Wind, data = airquality)
```

```{r}
airquality = transform(airquality, Month = factor(Month))
xyplot(Ozone ~ Wind | Month, data = airquality, layout = c(5,1))
```

Lattice behavior
- lattice graphic functions returns object of class trellis.
- print method for lattice do the work of plotting the data on graphics device. 


Lattice panel functions
- panel functions receive x and y coords along with arguments

```{r}
set.seed(100)
x = rnorm(100)
f = rep(0:1, each = 50)
y = x + f - f * x + rnorm(100, sd = .5)

f = factor(f, labels = c('G1','G2'))
xyplot(y ~ x | f, layout = c(2,1))
```

Custom panel function
```{r}
xyplot(y ~ x | f, panel  = function(x,y,...){
      panel.xyplot(x,y, ...) # first call default panel for xyplot
      panel.abline(h = median(y), lty = 2)
})
```


add a regression line
```{r}
xyplot(y ~ x | f, panel  = function(x,y,...){
      panel.xyplot(x,y, ...) # first call default panel for xyplot
      panel.lmline(x,y, col = 2)
})
```

## Ggplot2
- Implementation of grammar of graphics
- shorten distance from mind to page. 

### qplot
- quick plot basics
- dataframe to organize data
- aesthetics (size, shape, color) and geoms(points, lines)
- factors indicate subsets of data. 


```{r}
library(ggplot2)
str(mpg)
```

```{r}
qplot(displ, hwy, data = mpg, color = drv)
```


Adding geom
```{r}
qplot(displ, hwy, data = mpg, geom = c('point','smooth'))
```

```{r}
qplot(hwy, data = mpg, fill = drv)
```
Facets
```{r}
qplot(displ, hwy, data = mpg, facets = . ~ drv)
```


```{r}
qplot(hwy, data = mpg, facets = drv ~., binwidth = 2)
```

Components
- dataframe
- aesthetic mappings : how data are mapped to color, size, points
- geoms : geometric objects like points, lines, shapes.
- facets : conditional plots
- stats : statistical transformations like binning, quantiles, smoothing
- scales : how different variables are encoded
- coordinate system

artist palette model. 
Plots are buildup in layers
- plot data
- overlay a summary
- metadata and annotation

Annotations
- xlab(), ylab(), labs(), ggtitle()
- each of the geom has options to modify
- for things that only makes sense globally, use theme()
- theme_bw() : plain, theme_gray() : default

Notes on axis limits

```{r}
testdat = data.frame(x = 1:100, y = rnorm(100))
testdat[50,2] = 100

plot(testdat$x, testdat$y, type = 'l', ylim = c(-3,3))
```

Missing outlier
```{r}
g = ggplot(testdat, aes(x,y))
g + geom_line() + ylim(-3,3)
```

```{r}
g + geom_line() + coord_cartesian(ylim = c(-3,3))
```


## quiz


```{r}
library(nlme)
library(lattice)
xyplot(weight ~ Time | Diet, BodyWeight)
```

```{r}
library(datasets)
data(airquality)

qplot(Wind, Ozone, data = airquality)

airquality = transform(airquality, Month = factor(Month))
qplot(Wind, Ozone, data = airquality, facets = . ~ Month)

qplot(Wind, Ozone, data = airquality, facets = . ~ factor(Month))
```















